# Procesamiento Data Clases Chile

Capítulo para procesar la Data Conclat Chile con los esquemas de clases ya generados. Datos actualizados para el 26/03/2026.

## Preparación

```{r librerias, echo=FALSE, warning=FALSE, message=FALSE}
rm(list=ls())
options(scipen = 999)

# install.packages("pacman")
library(pacman)
p_load(
  # generacion docs 
  flextable,         # tablas personalizadas para word
  kableExtra,        # mejorar tablas
  knitr,             # generar reportes reproducibles
  officer,           # crear/modifcar documentos word o ppt
  tinytex,           # instalación mínima de LaTeX para generar PDF
  
  # importacion datos 
  haven,             # impotar/exportar datos spss, stata, sas con etiquetas
  readr,             # lectura rápida de csv
  readxl,            # leer excel 
  
  # manipulacion 
  forcats,           # manipulación de factores
  janitor,           # limpia etiquetas
  labelled,          # trabajar con variables etiquetadas
  purrr,             # programación funcional para listas y data frames
  sjlabelled,        # manejar etiquetas 
  stringr,           # manipular texto
  sjmisc,            # limpieza y descripción de datos
  
  # visualizacion
  ggplot2,           # visualización basado en gramática de gráficos
  ggrepel,           # etiquetas en gráficos sin superposición
  gridExtra,         # combinar gráficos en uno solo
  scales,            # controla cómo se ven los ejes en gráficos
  sjPlot,            # visualizar resultados estadísticos
  
  # analisis
  psych,             # análisis general, psicométrico
  survey,            # diseño survey encuestas complejas
  srvyr,             # diseño survey con tidy
  rlang,             # motor interno de tidyverse, para usar expresiones 
  tidyverse,
  statar             # conectar con el lenguaje stata

  # # conexion web
  # curl,              # conexiones http de bajo nivel: para solucionar el problema de la api del ine
  # httr               # cliente http para API
  )
```

### Cargar datos

```{r datos, message=FALSE, warning=FALSE}
data <- readRDS("output/data/data-clases-cl.rds")
```

## Procesamiento

### c02_rec

```{r}
frq(data$c02)

labs <- get_labels(data$c02, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c02_rec = if_else(
      c02 %in% c(-888, -999), NA_real_, as.numeric(c02)
    )
  ) %>% 
  set_variable_labels(c02_rec = "Preferred role of trade unions") %>%
  set_value_labels(c02_rec = labs_num)

frq(data$c02_rec)
```

### c03_01_rec

```{r}
frq(data$c03_01)

labs <- get_labels(data$c03_01, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c03_01_rec = if_else(
      c03_01 %in% c(-888, -999), NA_real_, as.numeric(c03_01)
    )
  ) %>% 
  set_variable_labels(c03_01_rec = "Level of agreement with: Companies have the right to replace striking workers") %>%
  set_value_labels(c03_01_rec = labs_num)

frq(data$c03_01_rec)
```

### c03_02_rec

```{r}
frq(data$c03_02)

labs <- get_labels(data$c03_02, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c03_02_rec = if_else(
      c03_02 %in% c(-888, -999), NA_real_, as.numeric(c03_02)
    )
  ) %>% 
  set_variable_labels(c03_02_rec = "Level of agreement with: Workers should not strike if it affects profits") %>%
  set_value_labels(c03_02_rec = labs_num)

frq(data$c03_02_rec)
```

### c05_01_rec

```{r}
frq(data$c05_01)

labs <- get_labels(data$c05_01, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c05_01_rec = if_else(
      c05_01 %in% c(-888, -999), NA_real_, as.numeric(c05_01)
    )
  ) %>% 
  set_variable_labels(c05_01_rec = "Level of agreement with: Large business owners have too much power in Chile") %>%
  set_value_labels(c05_01_rec = labs_num)

frq(data$c05_01_rec)
```

### c05_02_rec

```{r}
frq(data$c05_02)

labs <- get_labels(data$c05_02, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c05_02_rec = if_else(
      c05_02 %in% c(-888, -999), NA_real_, as.numeric(c05_02)
    )
  ) %>% 
  set_variable_labels(c05_02_rec = "Level of agreement with: A main reason for high inequality in Chile is that business owners pay workers too little") %>%
  set_value_labels(c05_02_rec = labs_num)

frq(data$c05_02_rec)
```

### c05_03_rec

```{r}
frq(data$c05_03)

labs <- get_labels(data$c05_03, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c05_03_rec = if_else(
      c05_03 %in% c(-888, -999), NA_real_, as.numeric(c05_03)
    )
  ) %>% 
  set_variable_labels(c05_03_rec = "Level of agreement with: High living standards of business owners result from paying workers too little") %>%
  set_value_labels(c05_03_rec = labs_num)

frq(data$c05_03_rec)
```

### c05_04_rec

```{r}
frq(data$c05_04)

labs <- get_labels(data$c05_04, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c05_04_rec = if_else(
      c05_04 %in% c(-888, -999), NA_real_, as.numeric(c05_04)
    )
  ) %>% 
  set_variable_labels(c05_04_rec = "Level of agreement with: Workers would live better if business owners had less power") %>%
  set_value_labels(c05_04_rec = labs_num)

frq(data$c05_04_rec)
```

### c05_05_rec

```{r}
frq(data$c05_05)

labs <- get_labels(data$c05_05, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c05_05_rec = if_else(
      c05_05 %in% c(-888, -999), NA_real_, as.numeric(c05_05)
    )
  ) %>% 
  set_variable_labels(c05_05_rec = "Level of agreement with: In Chile there are precarious and unstable jobs because this benefits business owners") %>%
  set_value_labels(c05_05_rec = labs_num)

frq(data$c05_05_rec)
```

### c07_01_rec

```{r}
frq(data$c07_01)

labs <- get_labels(data$c07_01, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c07_01_rec = if_else(
      c07_01 %in% c(-888, -999), NA_real_, as.numeric(c07_01)
    )
  ) %>% 
  set_variable_labels(c07_01_rec = "Level of agreement with: Increasing taxes on the wealthiest to finance social programs") %>%
  set_value_labels(c07_01_rec = labs_num)

frq(data$c07_01_rec)
```

### c07_03_rec

```{r}
frq(data$c07_03)

labs <- get_labels(data$c07_03, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c07_03_rec = if_else(
      c07_03 %in% c(-888, -999), NA_real_, as.numeric(c07_03)
    )
  ) %>% 
  set_variable_labels(c07_03_rec = "Level of agreement with: Implementing laws to give unions greater bargaining power") %>%
  set_value_labels(c07_03_rec = labs_num)

frq(data$c07_03_rec)
```

### c07_03_dummy

```{r}
frq(data$c07_03)

data <- data %>%
  mutate(
    c07_03_dummy = case_when(
      c07_03 %in% c(4, 5) ~ 1,
      c07_03 %in% c(1, 2, 3, -888, -999) ~ 0, 
      .default = NA_real_
    ))%>% 
  set_variable_labels(c07_03_dummy = "Level of agreement: Implement laws to give unions more bargaining power (1 = agree)") %>% 
  set_value_labels(c07_03_dummy = c(
    "Agree" = 1,
    "Disagree" = 0
  ))

frq(data$c07_03_dummy)
```

### c07_04_rec

```{r}
frq(data$c07_04)

labs <- get_labels(data$c07_04, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c07_04_rec = if_else(
      c07_04 %in% c(-888, -999), NA_real_, as.numeric(c07_04)
    )
  ) %>% 
  set_variable_labels(c07_04_rec = "Level of agreement with: Redistributing wealth from those with more resources to those with fewer") %>%
  set_value_labels(c07_04_rec = labs_num)

frq(data$c07_04_rec)
```

### c13_01_rec

```{r}
frq(data$c13_01)

labs <- get_labels(data$c13_01, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c13_01_rec = if_else(
      c13_01 %in% c(-888, -999), NA_real_, as.numeric(c13_01)
    )
  ) %>% 
  set_variable_labels(c13_01_rec = "Required level of change: Pension system") %>%
  set_value_labels(c13_01_rec = labs_num)

frq(data$c13_01_rec)
```

### c13_02_rec

```{r}
frq(data$c13_02)

labs <- get_labels(data$c13_02, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c13_02_rec = if_else(
      c13_02 %in% c(-888, -999), NA_real_, as.numeric(c13_02)
    )
  ) %>% 
  set_variable_labels(c13_02_rec = "Required level of change: Workers’ bargaining power vis-à-vis employers") %>%
  set_value_labels(c13_02_rec = labs_num)

frq(data$c13_02_rec)
```

### c13_03_rec

```{r}
frq(data$c13_03)

labs <- get_labels(data$c13_03, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c13_03_rec = if_else(
      c13_03 %in% c(-888, -999), NA_real_, as.numeric(c13_03)
    )
  ) %>% 
  set_variable_labels(c13_03_rec = "Required level of change: Economic model") %>%
  set_value_labels(c13_03_rec = labs_num)

frq(data$c13_03_rec)
```

### c13_04_rec

```{r}
frq(data$c13_04)

labs <- get_labels(data$c13_04, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c13_04_rec = if_else(
      c13_04 %in% c(-888, -999), NA_real_, as.numeric(c13_04)
    )
  ) %>% 
  set_variable_labels(c13_04_rec = "Required level of change: Income distribution") %>%
  set_value_labels(c13_04_rec = labs_num)

frq(data$c13_04_rec)
```

### c14_01_rec

```{r}
frq(data$c14_01)

labs <- get_labels(data$c14_01, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c14_01_rec = if_else(
      c14_01 %in% c(-888, -999), NA_real_, as.numeric(c14_01)
    )
  ) %>% 
  set_variable_labels(c14_01_rec = "Level of agreement with: Capitalism allows everyone to have opportunities to succeed") %>%
  set_value_labels(c14_01_rec = labs_num)

frq(data$c14_01_rec)
```

### c14_02_rec

```{r}
frq(data$c14_02)

labs <- get_labels(data$c14_02, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c14_02_rec = if_else(
      c14_02 %in% c(-888, -999), NA_real_, as.numeric(c14_02)
    )
  ) %>% 
  set_variable_labels(c14_02_rec = "Level of agreement with: Capitalism produces social injustices") %>%
  set_value_labels(c14_02_rec = labs_num)

frq(data$c14_02_rec)
```

### c14_03_rec

```{r}
frq(data$c14_03)

labs <- get_labels(data$c14_03, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c14_03_rec = if_else(
      c14_03 %in% c(-888, -999), NA_real_, as.numeric(c14_03)
    )
  ) %>% 
  set_variable_labels(c14_03_rec = "Level of agreement with: Under capitalism, only the rich benefit") %>%
  set_value_labels(c14_03_rec = labs_num)

frq(data$c14_03_rec)
```

### c14_04_rec

```{r}
frq(data$c14_04)

labs <- get_labels(data$c14_04, values = "n")
labs_clean <- labs[!names(labs) %in% c("-888", "-999")]
labs_num <- setNames(as.numeric(names(labs_clean)), labs_clean)

data <- data %>%
  mutate(
    c14_04_rec = if_else(
      c14_04 %in% c(-888, -999), NA_real_, as.numeric(c14_04)
    )
  ) %>% 
  set_variable_labels(c14_04_rec = "Level of agreement with: Capitalism is harmful to workers") %>%
  set_value_labels(c14_04_rec = labs_num)

frq(data$c14_04_rec)
```

### workingclass_id

```{r}
frq(data$d01)

data <- data %>% 
  mutate(
    workingclass_id = case_when(
      d01 %in% c(1, 2, 3) ~ 1, 
      d01 %in% c(4, 5, 6, -888, -999) ~ 0, 
      .default = NA_real_
    )
  ) %>% 
  set_variable_labels(workingclass_id = "Identification with the working class (1 = yes)") %>% 
  set_value_labels(workingclass_id = c(
    "Yes" = 1,
    "No" = 0
  ))

frq(data$workingclass_id)
```

### workingclass_people_id

```{r}
frq(data$d02_01)

data <- data %>% 
  mutate(
    workingclass_people_id = case_when(
      workingclass_id == 0 ~ 0,
      workingclass_id == 1 & d02_01 %in% c(-777, -888, -999) ~ 0,
      workingclass_id == 1 & d02_01 == 1 ~ 1, 
      workingclass_id == 1 & d02_01 == 2 ~ 2,
      workingclass_id == 1 & d02_01 == 3 ~ 3,
      workingclass_id == 1 & d02_01 == 4 ~ 4,
      workingclass_id == 1 & d02_01 == 5 ~ 5,
      .default = NA_real_
    )
  ) %>% 
  set_variable_labels(workingclass_people_id = "I identified with people from working class")
    
frq(data$workingclass_people_id)
```

### workingclass_id2

```{r}
frq(data$d03)

data <- data %>% 
  mutate(
    workingclass_id2 = case_when(
      d03 %in% c(4, 5, 6) ~ 1, 
      d03 %in% c(1, 2, 3, 7, -888, -999) ~ 0, 
      .default = NA_real_
    )
  ) %>% 
  set_variable_labels(workingclass_id2 = "Identification with the working class (1 = yes) (2)") %>% 
  set_value_labels(workingclass_id2 = c(
    "Yes" = 1,
    "No" = 0
  ))

frq(data$workingclass_id2)
```

## Controls

### female

```{r}
frq(data$e01)

data <- data %>%
  mutate(
    female = case_when(
      e01 == 1 ~ 0,
      e01 == 2 ~ 1
    )
  ) %>% 
  set_variable_labels(female = "Gender (1 = female)") %>% 
  set_value_labels(female = c(
    "Female" = 1, 
    "Male" = 0
  ))

frq(data$female)
```

### age

```{r}
descr(data$e02)

data <- data %>%
  mutate(
    age = if_else(
      e02 %in% c(-888, -999), NA_real_, as.numeric(e02)
    )
  ) %>% 
  set_variable_labels(age = "Age")

descr(data$age)
```

### sector

```{r}
frq(data$a07)

data <- data %>%
  mutate(
    sector=case_when(
      as.numeric(a07) %in% c(1, 2, 5, 6, 7, 9) ~ 1,
      as.numeric(a07) %in% c(3, 4, 8) ~ 0,
      .default = NA_real_
    )) %>% 
  set_variable_labels(sector = "Sector (1 = private)") %>% 
  set_value_labels(sector = c(
    "Private" = 1,
    "Public" = 0
  ))

frq(data$sector)
```

### rama

```{r,warning=FALSE,message=FALSE}
frq(data$a25_rec)

data <- data %>% 
  mutate(
    rama = a25_rec
  ) %>% 
  set_variable_labels(rama = "Economic Activity")

frq(data$rama)
```

### pol_id

```{r}
frq(data$c08)

data <- data %>%
  mutate(
    pol_id = factor(
      case_when(
        c08 %in% c(0, 1, 2, 3) ~ 1,
        c08 %in% c(4, 5, 6) ~ 2,
        c08 %in% c(7, 8, 9, 10) ~ 3,
        c08 %in% c(11, 12, -888, -999) ~ 4,
        .default = NA_real_
      ),
      levels = 1:4, 
      labels = c(
        "Left", 
        "Center", 
        "Right", 
        "Not identified"
        ) 
      )
    ) %>%
  set_variable_labels(pol_id = "Political identification") 

frq(data$pol_id)
```

### left_id

```{r}
frq(data$c08)

data <- data %>%
  mutate(
    left_id = if_else(
      c08 %in% c(0, 1, 2, 3), 1, 0
      )
  ) %>% 
  set_variable_labels(left_id = "Left identification (1 = yes") %>% 
  set_value_labels(left_id = c(
    "Yes" = 1,
    "No" = 0
  ))

frq(data$left_id)
```

## Guardar datos

```{r}
saveRDS(data, file="output/data/data-clases-cl.rds")
```
